---
title: 'Final Project 2: Covid 19 Data'
author: "Samantha Yow"
date: "2024-02-14"
output:
  pdf_document: default
  html_document: default
editor_options:
  chunk_output_type: console
---

This is the Final Project 2 for the course "Data Science as a Field" by Professor Jane Wall.  The first part is the demonstration by the Professor during the online class to show a Data Science Process.  The second part is a new analysis using the same data base by the author.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

During the project, I used the following library.

```{r library, echo=FALSE}
library(tidyverse)
library(lubridate)
library(tidyr)
library(ggplot2)

```

## Part 1 - By Instructor
The data files are from Github.  It consists of data about COVID-19 in different countries around the world and also different states of USA.  The date range of the data is from 2020 to 2023. They are in csv formate.  

```{r read_data, echo=TRUE}
url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"
file_names <-  c("time_series_covid19_confirmed_global.csv",
   "time_series_covid19_deaths_global.csv",
   "time_series_covid19_confirmed_US.csv" ,
   "time_series_covid19_deaths_US.csv")

urls <- str_c(url_in,file_names)

```

After we download the files, we read in the data to our project.

```{r import_data, message=FALSE}
global_cases <- read.csv(urls[1])
global_deaths <- read.csv(urls[2])
US_cases <- read.csv(urls[3])
US_deaths <- read.csv(urls[4])

```

The data files consist of a lot of useless data for our project, so we looked at the data and clean up the unwant part.  We also combine the data in different data frame into one dataframe for global data and one dataframe for states in the US.

```{r cleaning_data_global, echo=TRUE}
global_cases <- global_cases %>%
  pivot_longer(cols = -c("Province.State", "Country.Region","Lat","Long"),
               names_to = "date",
               values_to = "cases") %>%
  select(-c(Lat, Long))

global_cases[] <-lapply(global_cases, gsub, pattern='X', replacement='')


global_deaths <- global_deaths %>%
  pivot_longer(cols = -c("Province.State", "Country.Region","Lat","Long"),
               names_to = "date",
               values_to = "deaths") %>%
  select(-c(Lat, Long))

global_deaths[] <-lapply(global_deaths, gsub, pattern='X', replacement='')


global <- global_cases %>%
  full_join(global_deaths) %>%
  rename(Country_Region = Country.Region,
         Province_State = Province.State) %>%
  mutate(date = mdy(date))

global <- global %>%
  filter(cases>0)
  
```


```{r clean_data_US}
US_cases <- US_cases %>%
  pivot_longer(cols =-(UID:Combined_Key),
               names_to = "date",
               values_to = "cases") %>%
  filter(cases>=0)  %>%
  select(Admin2:cases) %>%
  select(-c(Lat, Long_))

US_cases[] <-lapply(US_cases, gsub, pattern = "X", replacement = "")

US_deaths <- US_deaths %>%
  pivot_longer(cols =-(UID:Population),
               names_to = "date",
               values_to = "deaths") %>%
  filter(deaths>=0)  %>%
  select(Admin2:deaths) %>%
  select(-c(Lat, Long_))

US_deaths[] <-lapply(US_deaths, gsub, pattern = "X", replacement = "")



US <- US_cases  %>%
  full_join(US_deaths)

```

```{r add_pop_global}

global <- global %>%
  unite("Combined_Key",
        c(Province_State, Country_Region),
        sep = ", ",
        na.rm = TRUE,
        remove = FALSE)


```

Here, we lookup for the populations and read in the file from Github for our analysis.

```{r lookup_global_pop}

uid_lookup_url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"

uid <- read.csv(uid_lookup_url) %>%
  select(-c(Lat, Long_, Combined_Key, code3, iso2, iso3, Admin2))

global <- global %>%
  left_join(uid, by=c("Province_State","Country_Region")) %>%
  select(-c(UID,FIPS)) %>%
  select(Province_State, Country_Region, date, cases, deaths, Population, Combined_Key)

global$cases <- as.double(global$cases)
global$deaths <- as.double(global$deaths)


```

```{r US_analysis}

US$cases <- as.double(US$cases)
US$deaths <- as.double(US$deaths)
US$Population <- as.double(US$Population)
US$date <- mdy(US$date)


US_by_state <- US %>%
  group_by(Province_State, Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths), Population = sum(Population)) %>%
  mutate(deaths_per_mill = deaths*1000000/Population) %>%
  select(Province_State, Country_Region, date, cases, deaths, deaths_per_mill, Population) %>%
  ungroup()

US_totals <- US_by_state %>%
  group_by(Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths), Population = sum(Population)) %>%
  mutate(deaths_per_mill = deaths*1000000/Population) %>%
  select(Country_Region, date, cases, deaths, deaths_per_mill, Population) %>%
  ungroup()


```

First, we plot the visualization for the US data.

```{r visualizations}

US_totals %>%
  filter(cases > 0) %>%
  ggplot(aes(x=date, y=cases)) +
  geom_line(aes(color = "cases")) +
  geom_point(aes(color = "cases")) +
  geom_line(aes(y=deaths, color = "deaths")) +
  geom_point(aes(y=deaths, color = "deaths")) +
  scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle=90)) +
  labs(title= "COVID19 in US", y=NULL)

state <- "New York"
US_by_state %>%
  filter(Province_State == state) %>%
  filter(cases > 0) %>%
  ggplot(aes(x=date, y=cases)) +
  geom_line(aes(color = "cases")) +
  geom_point(aes(color = "cases")) +
  geom_line(aes(y=deaths, color = "deaths")) +
  geom_point(aes(y=deaths, color = "deaths")) +
  scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle=90)) +
  labs(title= str_c("COVID19 in ",state , y=NULL))

```



```{r Analysis2}

US_by_state <- US_by_state %>%
  mutate(new_cases = cases-lag(cases),
         new_deaths = deaths-lag(deaths))


US_totals <- US_totals %>%
  mutate(new_cases = cases-lag(cases),
         new_deaths = deaths-lag(deaths))

```

Then, we transform the data of cases and deaths to new_cases and new_deaths.  Then plot the graphs with the new variables.


```{r visualization_new_cases}

US_totals %>%
  filter(cases > 0) %>%
  ggplot(aes(x=date, y=new_cases)) +
  geom_line(aes(color = "new_cases")) +
  geom_point(aes(color = "new_cases")) +
  geom_line(aes(y=new_deaths, color = "new_deaths")) +
  geom_point(aes(y=new_deaths, color = "new_deaths")) +
  scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle=90)) +
  labs(title= "COVID19 in US", y=NULL)

state <- "New York"
US_by_state %>%
  filter(Province_State == state) %>%
  filter(new_cases > 0) %>%
  ggplot(aes(x=date, y=new_cases)) +
  geom_line(aes(color = "new_cases")) +
  geom_point(aes(color = "new_cases")) +
  geom_line(aes(y=new_deaths, color = "new_deaths")) +
  geom_point(aes(y=new_deaths, color = "new_deaths")) +
  scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle=90)) +
  labs(title= str_c("COVID19 in ",state , y=NULL))

```

Then we clean up the data for the linear modelling.

```{r model}

US_state_totals <- US_by_state %>%
  group_by(Province_State) %>%
  summarize(deaths = max(deaths), cases = max(cases),
            population = max(Population),
            cases_per_thou = 1000*cases/population,
            deaths_per_thou = 1000*deaths/population) %>%
  filter(cases>0, population>0)

US_state_totals %>%
  slice_min(deaths_per_thou, n=10) %>%
  select(deaths_per_thou, cases_per_thou, everything())

US_state_totals %>%
  slice_max(deaths_per_thou, n=10) %>%
  select(deaths_per_thou, cases_per_thou, everything())


```

Last, the linear model comparing the deaths_per_thou and cases_per_thou is as follows

```{r linear_model}

mod <- lm(deaths_per_thou ~ cases_per_thou, data = US_state_totals)

US_tot_w_pred <- US_state_totals %>%
  mutate(pred = predict(mod))

US_tot_w_pred %>%
  ggplot() +
  geom_point(aes(x=cases_per_thou, y=deaths_per_thou), color ="blue") +
  geom_point(aes(x=cases_per_thou, y=pred), color ="red")

```

## Part 2 - by the author

In part 2, I would like to see if the temperature of the place will affect the infection rate or the death rate.  So, I find a database by "National Centers for Environmental Information" in this url link: https://www.ncei.noaa.gov/products/land-based-station/comparative-climatic-data
I download the data base and put it at the same directory of this file.


I calculated the Annual Average temperature for 50 states in USA and use the number as an indication of how ward or cold at the states.  I would like to see if the general temperature condition will affect the infection of COVID-19 or the number of deaths due to COVID-19.

```{r import_temp_data}

nrmavg <- read.csv("C:/Users/samny/OneDrive/Desktop/Notes of MSc/DS as a Field/Trial projects/nrmavg.csv")
  
View(nrmavg)

normal_temp <- nrmavg %>%
  group_by(Provience_State) %>%
  summarize(ANN = mean(ANN)) %>%
  select(Provience_State, ANN) %>%
  ungroup()

normal_temp <- rename(normal_temp, Province_State = Provience_State )

state_tot_w_temp <- US_state_totals %>%
  inner_join(normal_temp, by = "Province_State")
  
```

The two graphs below are a general plot to see if the temperature will affect the number of cases or the number of deaths.  It seems like the effect is more obvious in the deaths rates.  So, I further process the deaths rate data with a linear model.


```{r visual_temp}

state_tot_w_temp %>%
  ggplot(aes(x=ANN, y=cases_per_thou)) +
  geom_line(aes(color = "cases_per_thou")) +
  geom_point(aes(color = "cases_per_thou")) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle=90)) +
  labs(title= "COVID19 no. of cases relationship with Temperature", y=NULL)


state_tot_w_temp %>%
  ggplot(aes(x=ANN, y=deaths_per_thou)) +
  geom_line(aes(color = "deaths_per_thou")) +
  geom_point(aes(color = "deaths_per_thou")) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle=90)) +
  labs(title= "COVID19 no. of deaths relationship with Temperature", y=NULL)
```


```{r model_temp}

mod_temp_deaths <- lm(deaths_per_thou ~ ANN, data = state_tot_w_temp)

temp_deaths_pred <- state_tot_w_temp %>%
  mutate(pred = predict(mod_temp_deaths))

temp_deaths_pred %>%
  ggplot() +
  geom_point(aes(x=ANN, y=deaths_per_thou), color ="blue") +
  geom_point(aes(x=ANN, y=pred), color ="red")

summary(mod_temp_deaths)


```

After running the linear model, looking at the p-value, we can state that there is a positive relationship for the Average Annual Temperature and the deaths_per_thou.  For further analysis, I would suggest to do a linear relationship model for the cases_per_thou because there maybe a negative relationship between ANN and cases_per_thou.  We need to confirm that if this relationship is significant. Also we may consider to do a two factor linear model with (Deaths_per_thou ~ ANN + cases_per_thou).

For the bias in the data and this analysis, because this database is created in the US, the information in the US maybe more abundant and accurate.  And for my own bias, since I am leaving in another part of the world, I tend to search the US government data base for the average temperature and I am not very familiar with the other common used source of data in USA.