---
title: "NYPD Shooting Case"  
author: "Samantha Yow"  
date: "2024-1-28" 
output:
  html_document: default
  pdf_document: default
  word_document: default
editor_options: 
  chunk_output_type: console
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```  


### Project Purpose    

The project is an assignment of the course "Data Science As A Field".  This assignment is to complete all steps in the data science process in a reproducible manner. We use the NYPD Shooting Incident data we have been looking at during the course.

```{r run1, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(lubridate)
library(pacman)
```

### Upload and Cleaning of data    
First, we need to upload the data and take a look of it.  Then we remove all the rows with abnormal values.  After, we choose the columns we will use in our analysis.      
```{r run2, include=TRUE , echo=TRUE}
url_in <- "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD" 
nypd <- read_csv(url_in)  #read in the data

head(nypd)                #take a look of all the columns

Cases_completed <- nypd[complete.cases(nypd), ]   
#remove all the data point with empty field or abnormal values 

ShootingCases <- Cases_completed %>%   #select the columns we want to study 
    select(PERP_AGE_GROUP, PERP_SEX, VIC_AGE_GROUP)  

ShootingCases <- ShootingCases %>%
filter(!is.null(PERP_AGE_GROUP))
#remove all the data point with empty field or abnormal values 

ShootingCases <- ShootingCases %>%
filter(PERP_AGE_GROUP != "(null)")

ShootingCases <- ShootingCases %>%
filter(PERP_AGE_GROUP != "NA") %>%
filter(PERP_SEX != "NA") %>%
filter(VIC_AGE_GROUP != "NA")

#take a look of the data to make sure it is cleaned
head(ShootingCases)
summary(ShootingCases)
```

```{r visualprep, include=FALSE, echo=TRUE}

ShootingCases1 <- ShootingCases %>% count(PERP_AGE_GROUP, VIC_AGE_GROUP)
ShootingCases2 <- ShootingCases %>% count(PERP_SEX, VIC_AGE_GROUP)

summary(ShootingCases1)
summary(ShootingCases2)
```

I selected the VIC_AGE_GROUP as the dependent variable, and PERP_AGE_GROUP together with PERP_SEX as the independent variable. I want to know if the Victim's age is related to the Shooter's age and sex.

### Visualizaion
I created to charts to see if there is evidence that VIC_AGE_GROUP is related to PERP_AGE_GROUP and PERP_SEX seperately.

```{r visual, include=TRUE, echo=TRUE}
ggplot(data = ShootingCases1, mapping = aes(PERP_AGE_GROUP, VIC_AGE_GROUP))  + 
  geom_point(aes(size = n), color = "red")
ggplot(data = ShootingCases2, mapping = aes(PERP_SEX, VIC_AGE_GROUP))  +
  geom_point(aes(size = n), color = "blue")
```  

### Statistical Model  

Since all three variables are categorical data, I used the Poisson Distribution to run a regression to see if I can reasonbly predict the Victim's age group if I know the Shooter's age group and gender.

```{r model, include=TRUE, echo=TRUE}

#make sure the categorical data are well defined
age_levels <- c("<18","18-24","25-44","45-64","65+")
sex_levels <- c("M","F")

ShootingCasesM <- within(ShootingCases, {
  PERP_AGE_GROUP <- factor(PERP_AGE_GROUP, levels = age_levels)
  PERP_SEX <- factor(PERP_SEX, levels = sex_levels)
  VIC_AGE_GROUP <- factor(VIC_AGE_GROUP, levels = age_levels)
})

ShootingCasesM$VIC_AGE_GROUP <- unclass(ShootingCasesM$VIC_AGE_GROUP)
#change the victim age group to 1 represent ""<18",2 represent "18-24",3 represent #"25-44" etc.

ShootingCasesM <- ShootingCasesM %>% #further cleaning
filter(VIC_AGE_GROUP != "NA")

summary(ShootingCasesM)

head(ShootingCasesM)

regresult <- glm(formula = VIC_AGE_GROUP ~ PERP_AGE_GROUP + PERP_SEX , family = "poisson", data = ShootingCasesM)  #apply the model

summary(regresult)
```
The outcome of the Generalized Linear Model is as above.  We can see that the age groups of Shooters and Victims are related.  However, we can't see relation of Shooter's gender to the Victim's age group.

Then I tried 3 data point to see the result of predication.

```{r predict, echo=TRUE, warning=FALSE, include=TRUE}
newdata <- data_frame(PERP_AGE_GROUP = c("25-44", "65+", "18-24"),
                     PERP_SEX = c("M", "M", "F"))

newdata

newresult <- predict(regresult, newdata, level = 0.95, interval="confidence", type = "response")

newresult
```

### Bias during the Data Analysis Process

I must admit that I am not very strong in statistics modeling and I am not familiar with the Poisson Distribution. So, I was looking up how to use the model at the same time when I run the analysis.  There maybe something I missed during the process.  After taking this class and finished this project, I found my weakness in statistics.  Therefore, I decided to first finished the "Statistics Pathway" courses before I move on to other courses.  I finished the "Programming Pathway" alright and found it useful.  Hopefully, the "Statistics Pathway" will help me build up my skills and prepare me for the other courses.